---
title: "QC"
output: html_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load required R packages

**Source: Mixed: software implementation around the Nestorowa, Lun, and Brennecke methods.**

- `data.table` reads the large tab-delimited count table.
- `AnnotationDbi` and `org.Mm.eg.db` provide mouse annotation for gene types and symbols.Nestorowa originally used Ensembl release 81. However, I was not able to get 2016-2017 versions.
- `set.seed(100)` in case of any stochastic process.

```{r}
suppressPackageStartupMessages({
    library(data.table)
    library(AnnotationDbi)
    library(org.Mm.eg.db)
})

set.seed(100)
```

# 1. Nestorowa et al. — cell filtering and QC

Nestorowa first describes read assignment and then states the four requirements a cell must satisfy before any normalization is performed. This section therefore contains **only data preparation and cell filtering**.

## 1.1 Point R to the downloaded raw count table

The paper states that reads were aligned with G-SNAP, assigned to **Ensembl release 81 genes using HTSeq**, and deposited in GEO with id: GSE81682.

```{r}
counts_path <- "nestorowa_singlecell_raw_data_taken_from_original_paper_GSE.txt"
```

## 1.2 Read the HTSeq table into a numeric gene/feature × cell matrix

- `fread()` loads the tab-delimited file.
- The first column contains feature IDs; the remaining columns are cells.
- The feature IDs are moved to `rownames(raw_counts)`.
- The matrix is forced to numeric.
- `stopifnot()` checks that feature IDs are unique, no `NA` values were introduced, and no negative counts exist.
- The final `cat()` statements report the imported dimensions before any filtering.

```{r}
raw_df <- fread(
    counts_path,
    sep = "\t",
    data.table = FALSE,
    check.names = FALSE
)

feature_id <- as.character(raw_df[[1]])

raw_counts <- as.matrix(
    raw_df[, -1, drop = FALSE]
)

storage.mode(raw_counts) <- "numeric"
rownames(raw_counts) <- feature_id

stopifnot(!anyDuplicated(rownames(raw_counts)))
stopifnot(!anyNA(raw_counts))
stopifnot(all(raw_counts >= 0))

cat("Raw rows:", nrow(raw_counts), "\n")
cat("Raw cells:", ncol(raw_counts), "\n")
```

## 1.3 Identify endogenous genes, ERCC spike-ins, and HTSeq bookkeeping rows

The raw GEO table contains more than genes. Rows beginning `__` are HTSeq summary categories such as `__not_aligned`, not genes.

- `^ENSMUSG` identifies mouse Ensembl gene rows.
- `^ERCC[-_]` identifies ERCC spike-ins.
- `^__` identifies HTSeq summary rows.
- `is_unknown` is a safety check that stops the analysis if an unexpected row type appears.

This separation is necessary because HTSeq summary rows are useful for reconstructing QC totals but must never enter normalization or gene-variability analysis.

```{r}
# These patterns describe the row types present in the downloaded HTSeq table.
ENDOGENOUS_PATTERN <- "^ENSMUSG"
ERCC_PATTERN <- "^ERCC[-_]"
HTSEQ_PATTERN <- "^__"

ids <- rownames(raw_counts)

is_endogenous <- grepl(ENDOGENOUS_PATTERN, ids)
is_ercc <- grepl(ERCC_PATTERN, ids, ignore.case = TRUE)
is_htseq <- grepl(HTSEQ_PATTERN, ids)

is_unknown <- !(is_endogenous | is_ercc | is_htseq)

print(
    data.frame(
        class = c("ENSMUSG", "ERCC", "HTSeq summary", "unknown"),
        n = c(
            sum(is_endogenous),
            sum(is_ercc),
            sum(is_htseq),
            sum(is_unknown)
        )
    )
)

cat("\nHTSeq rows:\n")
print(ids[is_htseq])

if (any(is_unknown)) {
    stop(
        "Unexpected rows found: ",
        paste(ids[is_unknown], collapse = ", ")
    )
}
```

## 1.4 Define which Ensembl rows are mitochondrial

**Source: Nestorowa supplies the `<10% mitochondrial reads` rule**

Nestorowa does not provide R code or the exact mitochondrial-ID vector used in their analysis. The code therefore constructs the 37 GRCm38 mitochondrial Ensembl genes explicitly.

`is_mito_full` marks mitochondrial endogenous rows and `is_nuclear_full` defines the remaining endogenous genes as nuclear. Missing mitochondrial IDs are printed so the calculation can be audited.

For the code below I got help from GPT. 

```{r}
grcm38_mito_ensembl <- c(
    paste0("ENSMUSG000000", 64336:64361),
    paste0("ENSMUSG000000", 64363:64372),
    "ENSMUSG00000065947"
)

stopifnot(length(grcm38_mito_ensembl) == 37)

is_mito_full <- is_endogenous & ids %in% grcm38_mito_ensembl
is_nuclear_full <- is_endogenous & !is_mito_full

cat("Mitochondrial genes expected:", length(grcm38_mito_ensembl), "\n")
cat("Mitochondrial genes present in raw matrix:", sum(is_mito_full), "\n")

missing_mito <- setdiff(
    grcm38_mito_ensembl,
    ids
)

if (length(missing_mito) > 0) {
    cat("\nMitochondrial Ensembl IDs absent from the raw matrix:\n")
    print(missing_mito)
}

cat("\nMitochondrial rows actually used:\n")
print(ids[is_mito_full])
```

## 1.5 Separate endogenous counts, ERCC counts, and HTSeq summary counts


The mixed table is split into three matrices because each row type has a different role:

- `endogenous_counts` — biological mouse-gene expression;
- `ercc_counts` — external technical controls;
- `htseq_counts` — mapping/counting bookkeeping used only for QC reconstruction.

No cells are removed yet.

```{r}
endogenous_counts <- raw_counts[
    is_endogenous,
    ,
    drop = FALSE
]

ercc_counts <- raw_counts[
    is_ercc,
    ,
    drop = FALSE
]

htseq_counts <- raw_counts[
    is_htseq,
    ,
    drop = FALSE
]

cat(
    "Endogenous:", nrow(endogenous_counts),
    "\nERCC:", nrow(ercc_counts),
    "\nHTSeq summary:", nrow(htseq_counts),
    "\n"
)
```

## 1.6 Introduce the four Nestorowa QC thresholds

**Source: Nestorowa et al. (2016).**

These values are defined here, immediately before the QC quantities are calculated:

- at least **200,000 reads mapping to nuclear genes**;
- at least **4,000 detected genes**;
- **<10%** of mapped reads mapping to mitochondrial genes;
- **<50%** of mapped reads mapping to ERCC spike-ins.

`NESTOROWA_REPORTED_CELLS <- 1656` stores the published post-QC total only for comparison; it is not used to force the filtering result.

```{r}
# Nestorowa et al. (2016) cell-QC thresholds.
MIN_NUCLEAR_READS <- 200000
MIN_DETECTED_GENES <- 4000
MAX_MITO_PCT <- 10
MAX_ERCC_PCT <- 50

# Published number of cells passing QC in Nestorowa et al.
NESTOROWA_REPORTED_CELLS <- 1656
```

## 1.7 Reconstruct the QC quantities for every cell

The function `get_htseq_row()` retrieves one HTSeq summary category across all cells.

The code then constructs:

- `raw_total_reads` — the sum of every HTSeq-reported row;
- `mapped_reads` — reconstructed as total records minus `__not_aligned`;
- `nuclear_reads` — reads assigned to nuclear Ensembl genes;
- `mitochondrial_reads` — reads assigned to the mitochondrial set;
- `ercc_reads` — reads assigned to ERCC controls;
- `assigned_feature_reads` — reads assigned to endogenous genes or ERCCs.

Finally:

- `mito_pct_mapped = 100 × mitochondrial_reads / mapped_reads`;
- `ercc_pct_mapped = 100 × ercc_reads / mapped_reads`.

The resulting `qc` data frame stores all cell-level QC quantities before filtering.

```{r}
get_htseq_row <- function(row_name) {

    if (row_name %in% rownames(htseq_counts)) {
        return(
            as.numeric(
                htseq_counts[row_name, , drop = TRUE]
            )
        )
    }

    rep(0, ncol(raw_counts))
}

raw_total_reads <- colSums(raw_counts)

not_aligned <- get_htseq_row("__not_aligned")

mapped_reads <- raw_total_reads - not_aligned

nuclear_reads <- colSums(
    raw_counts[
        is_nuclear_full,
        ,
        drop = FALSE
    ]
)

mitochondrial_reads <- colSums(
    raw_counts[
        is_mito_full,
        ,
        drop = FALSE
    ]
)

ercc_reads <- colSums(ercc_counts)

assigned_feature_reads <- (
    colSums(endogenous_counts) +
    ercc_reads
)

# Get gene type for each endogenous Ensembl gene
gene_type <- AnnotationDbi::mapIds(
    org.Mm.eg.db,
    keys = ids[is_endogenous],
    keytype = "ENSEMBL",
    column = "GENETYPE",
    multiVals = "first"
)

is_protein_coding_endogenous <- gene_type == "protein-coding"
is_protein_coding_endogenous[is.na(is_protein_coding_endogenous)] <- FALSE

cat(
    "Protein-coding endogenous genes:",
    sum(is_protein_coding_endogenous),
    "\n"
)

detected_genes <- colSums(
    endogenous_counts[
        is_protein_coding_endogenous,
        ,
        drop = FALSE
    ] > 0
)

mito_pct_mapped <- (
    100 * mitochondrial_reads / mapped_reads
)

ercc_pct_mapped <- (
    100 * ercc_reads / mapped_reads
)

qc <- data.frame(
    cell = colnames(raw_counts),
    raw_total_reads = raw_total_reads,
    mapped_reads = mapped_reads,
    assigned_feature_reads = assigned_feature_reads,
    nuclear_reads = nuclear_reads,
    detected_genes = detected_genes,
    mitochondrial_reads = mitochondrial_reads,
    ercc_reads = ercc_reads,
    mito_pct_mapped = mito_pct_mapped,
    ercc_pct_mapped = ercc_pct_mapped,
    stringsAsFactors = FALSE
)

summary(
    qc[, c(
        "mapped_reads",
        "nuclear_reads",
        "detected_genes",
        "mito_pct_mapped",
        "ercc_pct_mapped"
    )]
)
```

## 1.8 Apply the four Nestorowa cell filters

**Source: Nestorowa et al. (2016)**

Each logical column corresponds directly to one published QC rule:

- `pass_nuclear` — ≥200,000 nuclear-gene reads;
- `pass_genes` — ≥4,000 detected genes;
- `pass_mito` — <10% mitochondrial reads;
- `pass_ercc` — <50% ERCC reads.

`pass_all` requires all four conditions simultaneously. The code reports the reconstructed surviving-cell count and compares it with the paper's **1,656** without altering any threshold to make the numbers match.

```{r}
qc$pass_nuclear <- (
    qc$nuclear_reads >= MIN_NUCLEAR_READS
)

qc$pass_genes <- (
    qc$detected_genes >= MIN_DETECTED_GENES
)

qc$pass_mito <- (
    qc$mito_pct_mapped < MAX_MITO_PCT
)

qc$pass_ercc <- (
    qc$ercc_pct_mapped < MAX_ERCC_PCT
)

qc$pass_all <- with(
    qc,
    pass_nuclear &
    pass_genes &
    pass_mito &
    pass_ercc
)

cat("Input cells:", nrow(qc), "\n")
cat("Post-QC cells:", sum(qc$pass_all), "\n")
cat("Paper:", NESTOROWA_REPORTED_CELLS, "\n")
cat("Difference:", sum(qc$pass_all) - NESTOROWA_REPORTED_CELLS, "\n\n")

cat("Fail nuclear:", sum(!qc$pass_nuclear), "\n")
cat("Fail genes:", sum(!qc$pass_genes), "\n")
cat("Fail mito:", sum(!qc$pass_mito), "\n")
cat("Fail ERCC:", sum(!qc$pass_ercc), "\n")
```

## 1.9 Inspect overlapping reasons for QC failure

A cell can fail more than one Nestorowa criterion. This cell builds a label such as `nuclear;genes;` for each cell and tabulates the combinations. It is only a diagnostic and does not change which cells pass.

```{r}
fail_pattern <- paste0(
    ifelse(qc$pass_nuclear, "", "nuclear;"),
    ifelse(qc$pass_genes, "", "genes;"),
    ifelse(qc$pass_mito, "", "mito;"),
    ifelse(qc$pass_ercc, "", "ERCC;")
)

fail_pattern[fail_pattern == ""] <- "PASS"

sort(
    table(fail_pattern),
    decreasing = TRUE
)
```

## 1.10 Create the QC-passing endogenous and ERCC matrices

- `keep_cells` contains cell names with `pass_all == TRUE`.
- `endogenous_qc` and `ercc_qc` retain exactly those same cells.
- Features that are zero in every retained cell are removed because they contain no information for normalization or variability analysis in future.
- The `stopifnot()` calls verify that no HTSeq `__...` summary rows have leaked into either expression matrix.

At this point **cell filtering is complete**. 

```{r}
keep_cells <- qc$cell[
    qc$pass_all
]

endogenous_qc <- endogenous_counts[
    ,
    keep_cells,
    drop = FALSE
]

ercc_qc <- ercc_counts[
    ,
    keep_cells,
    drop = FALSE
]

# Remove features that are zero in every retained cell.
endogenous_qc <- endogenous_qc[
    rowSums(endogenous_qc) > 0,
    ,
    drop = FALSE
]

ercc_qc <- ercc_qc[
    rowSums(ercc_qc) > 0,
    ,
    drop = FALSE
]

stopifnot(
    !any(grepl("^__", rownames(endogenous_qc))),
    !any(grepl("^__", rownames(ercc_qc)))
)

cat("Cells:", ncol(endogenous_qc), "\n")
cat("Endogenous genes:", nrow(endogenous_qc), "\n")
cat("ERCCs:", nrow(ercc_qc), "\n")
```

## 1.11 Compare cells between Chevalier et al. and this QC

```{r}
Chevalier_raw_data <- fread("nestorowa_singlecell_raw_data_taken_from_bonesis_paper.tsv",data.table = FALSE)

rownames(Chevalier_raw_data) <- Chevalier_raw_data[[1]]

Chevalier_raw_data <- Chevalier_raw_data[, -1,drop = FALSE]

cat("Chevalier et al. Cell Count:", ncol(Chevalier_raw_data), "\n")
cat("Intersect with this QC", length(intersect(colnames(Chevalier_raw_data),keep_cells)), "\n")
```

